---
title: "Genomic SEM: Sex-Stratified ARD Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  data: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table)
library(ggplot2)
data <- params$data
config <- params$config
```

## Overview

Pipeline: `r config$project$name`
Run date: `r Sys.Date()`
h2 Z-threshold: `r config$ldsc$h2_z_threshold`

## Heritability Estimates

### Male

```{r h2-male}
if (!is.null(data$male$h2)) {
  knitr::kable(data$male$h2, digits = 4, caption = "Male h2 estimates (observed scale)")
}
```

### Female

```{r h2-female}
if (!is.null(data$female$h2)) {
  knitr::kable(data$female$h2, digits = 4, caption = "Female h2 estimates (observed scale)")
}
```

## Exploratory Factor Analysis

### Male

```{r efa-male}
if (!is.null(data$male$efa_loadings)) {
  knitr::kable(data$male$efa_loadings, digits = 3, caption = "Male EFA loadings")
}
```

### Female

```{r efa-female}
if (!is.null(data$female$efa_loadings)) {
  knitr::kable(data$female$efa_loadings, digits = 3, caption = "Female EFA loadings")
}
```

## Confirmatory Factor Analysis

### Model Fit Comparison

```{r cfa-fit}
fit_all <- data.table()
for (sex in c("male", "female")) {
  if (!is.null(data[[sex]]$cfa_fit)) {
    tmp <- copy(data[[sex]]$cfa_fit)
    tmp[, sex := sex]
    fit_all <- rbind(fit_all, tmp, fill = TRUE)
  }
}
if (nrow(fit_all) > 0) {
  knitr::kable(fit_all, digits = 3, caption = "CFA model fit indices")
}
```

## Sex Comparison

```{r comparison}
if (!is.null(data$comparison) && nrow(data$comparison) > 0) {
  knitr::kable(data$comparison, digits = 3, caption = "Factor loading sex differences (unstandardized)")
} else {
  cat("No sex comparison data available.")
}
```

## Factor GWAS Outputs

```{r factor-gwas}
out_dir <- config$paths$output_dir
vcf_paths <- file.path(out_dir, c("male", "female"), "gwas",
                       c("male_chapter_gsem.vcf.gz", "female_chapter_gsem.vcf.gz"))
vcf_status <- data.frame(
  sex = c("male", "female"),
  path = vcf_paths,
  exists = file.exists(vcf_paths),
  size_mb = ifelse(file.exists(vcf_paths),
                    round(file.size(vcf_paths) / 1024^2, 2),
                    NA_real_)
)
knitr::kable(vcf_status,
             caption = "GWAS-VCF outputs: one file per sex, each containing one chapter-factor 'sample'")
```

## Scale and Units

The input data are Neale UKBB round-2 GWAS sumstats from **linear regression on the 0/1 phenotype** (not logistic). Each per-trait BETA/SE is converted to log-odds at munge time via

$$\beta_{\log\text{OR}} \;=\; \beta_{\text{linear}} \,/\, [K\,(1-K)], \qquad K = n_{\text{cases}} \,/\, (n_{\text{cases}} + n_{\text{controls}}).$$

Because Z = β/SE is invariant under this per-trait scaling, **LDSC heritability values are identical to what would be obtained on the raw linear-probability scale**, but now interpretable as h² on the log-odds (≈liability) scale. Downstream `GenomicSEM::sumstats()` is called with `se.logit=TRUE, OLS=FALSE, linprob=FALSE` (data already log-odds; do not double-convert), and `GenomicSEM::userGWAS()` produces factor BETAs as log(OR) per allele copy of ALT. The two output VCFs carry these log-odds effects in the `ES` FORMAT field per GWAS-VCF spec; see header lines `##gsem_scale=log_odds_ratio` and `##gsem_conversion=...`.

## Methods Note

- Effective sample size: Neff = 4 * n_cases * n_controls / (n_cases + n_controls).
- LDSC free intercept used to account for within-sex sample overlap.
- EFA and CFA applied to the same genetic covariance matrix (double-dipping caveat applies).
- Sex differences assessed via Wald Z-test on unstandardized loading differences from independent per-sex fits.
- Per-SNP factor GWAS: `GenomicSEM::userGWAS()` regresses each ICD-10 chapter latent factor on SNP, run independently per sex on the aligned 1000G EUR reference.
- Output format: [GWAS-VCF (MRC-IEU spec)](https://github.com/MRCIEU/gwas-vcf-specification), one file per sex, FORMAT fields ES (log OR), SE, LP (−log10 P), AF, SS.
