---
title: "Genomic SEM: Sex-Stratified ARD Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  data: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table)
library(ggplot2)
data <- params$data
config <- params$config
```

## Overview

Pipeline: `r config$project$name`
Run date: `r Sys.Date()`
h2 Z-threshold: `r config$ldsc$h2_z_threshold`

## Heritability Estimates

### Male

```{r h2-male}
if (!is.null(data$male$h2)) {
  knitr::kable(data$male$h2, digits = 4, caption = "Male h2 estimates (log-odds scale; ≈ liability for low-K traits — see Scale and Units)")
}
```

### Female

```{r h2-female}
if (!is.null(data$female$h2)) {
  knitr::kable(data$female$h2, digits = 4, caption = "Female h2 estimates (log-odds scale; ≈ liability for low-K traits — see Scale and Units)")
}
```

## Exploratory Factor Analysis

### Male

```{r efa-male}
if (!is.null(data$male$efa_loadings)) {
  knitr::kable(data$male$efa_loadings, digits = 3, caption = "Male EFA loadings")
}
```

### Female

```{r efa-female}
if (!is.null(data$female$efa_loadings)) {
  knitr::kable(data$female$efa_loadings, digits = 3, caption = "Female EFA loadings")
}
```

## Genetic-Correlation Heatmaps

```{r rg-heatmaps, fig.show='hold', out.width="49%"}
out_dir <- config$paths$output_dir
for (sex in c("male", "female")) {
  p <- file.path(out_dir, sex, "plots", sprintf("%s_rg_heatmap.png", sex))
  if (file.exists(p)) knitr::include_graphics(p)
}
```

## Confirmatory Factor Analysis

### Model Fit Comparison

```{r cfa-fit}
fit_all <- data.table()
for (sex in c("male", "female")) {
  if (!is.null(data[[sex]]$cfa_fit)) {
    tmp <- copy(data[[sex]]$cfa_fit)
    tmp[, sex := sex]
    fit_all <- rbind(fit_all, tmp, fill = TRUE)
  }
}
if (nrow(fit_all) > 0) {
  knitr::kable(fit_all, digits = 3, caption = "CFA model fit indices")
}
```

### Path Diagrams

```{r path-diagrams, fig.show='hold', out.width="49%"}
for (sex in c("male", "female")) {
  p <- file.path(out_dir, sex, "plots", sprintf("%s_path_diagram.png", sex))
  if (file.exists(p)) knitr::include_graphics(p)
}
```

### Loadings

```{r loadings-bar, fig.show='hold', out.width="49%"}
for (sex in c("male", "female")) {
  p <- file.path(out_dir, sex, "plots", sprintf("%s_loadings_bar.png", sex))
  if (file.exists(p)) knitr::include_graphics(p)
}
```

## Sex Comparison

```{r comparison}
if (!is.null(data$comparison) && nrow(data$comparison) > 0) {
  knitr::kable(data$comparison, digits = 3, caption = "Factor loading sex differences (unstandardized)")
} else {
  cat("No sex comparison data available.")
}
```

```{r sex-diff-forest, fig.align='center', out.width="80%"}
p <- file.path(out_dir, "comparison", "sex_diff_forest.png")
if (file.exists(p)) knitr::include_graphics(p)
```

## Factor GWAS

### Per-Factor Summary

```{r factor-summary}
fs <- data.table()
for (sex in c("male", "female")) {
  p <- file.path(out_dir, sex, "gwas", sprintf("%s_factor_summary.csv", sex))
  if (file.exists(p)) fs <- rbind(fs, fread(p), fill = TRUE)
}
if (nrow(fs) > 0) knitr::kable(fs, digits = 3,
                                caption = "Per-(sex × factor) GWAS summary")
```

### Manhattan + QQ plots

```{r manhattan-qq, fig.show='hold', out.width="49%"}
for (sex in c("male", "female")) {
  plot_dir <- file.path(out_dir, sex, "plots")
  if (!dir.exists(plot_dir)) next
  pngs <- list.files(plot_dir, pattern = "_(manhattan|qq)\\.png$", full.names = TRUE)
  for (p in pngs) knitr::include_graphics(p)
}
```

### Miami plots (male vs female per factor)

```{r miami, fig.show='hold', out.width="95%"}
miami_dir <- file.path(out_dir, "comparison")
if (dir.exists(miami_dir)) {
  pngs <- list.files(miami_dir, pattern = "^miami_.*\\.png$", full.names = TRUE)
  for (p in pngs) knitr::include_graphics(p)
}
```

### Top SNPs per factor

```{r top-snps}
for (sex in c("male", "female")) {
  p <- file.path(out_dir, sex, "gwas", sprintf("%s_top_snps.csv", sex))
  if (file.exists(p)) {
    tdt <- fread(p)
    if (nrow(tdt) > 0) {
      cat(sprintf("\n#### %s\n\n", toupper(sex)))
      print(knitr::kable(tdt, digits = 4))
    }
  }
}
```

## Factor GWAS Outputs

```{r factor-gwas-vcfs}
vcf_paths <- c(
  file.path(out_dir, "male",   "gwas", "male_chapter_gsem.vcf.gz"),
  file.path(out_dir, "male",   "gwas", "male_chapter_gsem.vcf.gz.tbi"),
  file.path(out_dir, "female", "gwas", "female_chapter_gsem.vcf.gz"),
  file.path(out_dir, "female", "gwas", "female_chapter_gsem.vcf.gz.tbi")
)
vcf_status <- data.frame(
  file = basename(vcf_paths),
  path = vcf_paths,
  exists = file.exists(vcf_paths),
  size_mb = ifelse(file.exists(vcf_paths),
                    round(file.size(vcf_paths) / 1024^2, 2),
                    NA_real_)
)
knitr::kable(vcf_status,
             caption = "GWAS-VCF outputs (bgzip + tabix-indexed): one .vcf.gz + .tbi pair per sex. Each VCF contains one chapter-factor 'sample'. Q_SNP heterogeneity is exposed per-factor as the HP FORMAT field; no variant-level Q_SNP_HET FILTER is applied here — filter downstream on HP per your analysis.")
```

## Scale and Units

The input data are Neale UKBB round-2 GWAS sumstats from **linear regression on the 0/1 phenotype** (not logistic). Each per-trait BETA/SE is converted to log-odds at munge time via

$$\beta_{\log\text{OR}} \;=\; \beta_{\text{linear}} \,/\, [K\,(1-K)], \qquad K = n_{\text{cases}} \,/\, (n_{\text{cases}} + n_{\text{controls}}).$$

Because Z = β/SE is invariant under this per-trait scaling, **LDSC heritability values are identical to what would be obtained on the raw linear-probability scale**, but now interpretable as h² on the log-odds (≈liability) scale. Downstream `GenomicSEM::sumstats()` is called with `se.logit=TRUE, OLS=FALSE, linprob=FALSE` (data already log-odds; do not double-convert), and `GenomicSEM::userGWAS()` produces factor BETAs as log(OR) per allele copy of ALT. The two output VCFs carry these log-odds effects in the `ES` FORMAT field per GWAS-VCF spec; see header lines `##gsem_scale=log_odds_ratio` and `##gsem_conversion=...`.

## Methods Note

- Effective sample size: Neff = 4 * n_cases * n_controls / (n_cases + n_controls).
- LDSC free intercept used to account for within-sex sample overlap. Population prevalence is intentionally set to NA (no liability-scale rescaling) — combined with the linear→log(OR) conversion at munge time, h² is reported on the log-odds scale, which approximates the liability-scale value for low-K traits without requiring an externally-provided population prevalence.
- EFA and CFA applied to the same genetic covariance matrix (double-dipping caveat applies).
- Sex differences assessed via Wald Z-test on unstandardised loading differences from independent per-sex fits. **Factor indicators may differ across sexes** when the h² Z-threshold filter retains different traits per sex — the sex-difference comparison only Wald-tests loadings on the **shared** indicators, but a chapter factor with indicator set {A,B} in male versus {A,B,C,D} in female is technically a different construct in each sex; this is methodologically defensible (sex-stratified factor models) but worth noting when interpreting the comparison table.
- Per-SNP factor GWAS: `GenomicSEM::userGWAS(..., std.lv = TRUE)` regresses each ICD-10 chapter latent factor on SNP, run independently per sex on the aligned 1000G EUR reference. With `std.lv = TRUE`, each factor's variance is fixed to 1 and SNP betas are reported in per-SD-of-factor units (comparable across factors).
- PSD smoothing: when the genetic-covariance matrix S fails the PSD check after the h²-Z filter, `Matrix::nearPD()` is applied to S only. The sampling-covariance matrix V is left unchanged, which means downstream sandwich SEs in `userGWAS` may be slightly biased on runs that triggered smoothing. The pipeline logs both the Frobenius-norm delta of the smoothing and the V-mismatch caveat.
- Output format: [GWAS-VCF (MRC-IEU spec)](https://github.com/MRCIEU/gwas-vcf-specification), one bgzipped + tabix-indexed file per sex (`.vcf.gz` + `.vcf.gz.tbi`). FORMAT fields: ES (log OR per ALT copy), SE, LP (−log10 P), AF, SS, HET (Q_SNP χ²), HP (Q_SNP p-value), SF (per-factor SEM-convergence flag: 1 = userGWAS did not converge at this SNP for this factor, 0 = converged).
- **Q_SNP heterogeneity** is exposed as the per-factor FORMAT field `HP`. We deliberately do **not** emit a variant-level FILTER for Q_SNP_HET — a SNP that is heterogeneous on one chapter factor is still informative for the others. Downstream consumers should filter on `HP` per their needs (the GenomicSEM wiki recommends `HP < 5e-8` plus LD pruning).
- **λ_GC reported in the VCF header (`##gsem_lambda_gc_<factor>`) and `factor_summary.csv` is the raw value** computed across all per-(SNP, factor) P-values, not the post-Q_SNP-filtered value some downstream papers report. To get the filtered λ_GC, recompute on the subset with `HP >= 5e-8`.
- Path diagrams are emitted as Graphviz `.dot` files (auto-rendered to PNG when `dot` is on PATH); embed in a paper by either committing the `.dot` and rendering with `dot -Tpdf`, or using the included PNG.
- Manhattan and Miami plots downsample non-significant SNPs (default: keep all SNPs with `P < 1e-3`, plus a random 10% of the rest) to make rendering tractable at 1M-SNP scale. The downsampling is purely cosmetic — every significant signal is retained at full density. Disable via `plots.manhattan_bg_keep_frac: 1.0` in `config/pipeline.yaml`.
