---
title: "Genomic SEM: Sex-Stratified ARD Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  data: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table)
library(ggplot2)
data <- params$data
config <- params$config
```

## Overview

Pipeline: `r config$project$name`
Run date: `r Sys.Date()`
h2 Z-threshold: `r config$ldsc$h2_z_threshold`

## Heritability Estimates

### Male

```{r h2-male}
if (!is.null(data$male$h2)) {
  knitr::kable(data$male$h2, digits = 4, caption = "Male h2 estimates (observed scale)")
}
```

### Female

```{r h2-female}
if (!is.null(data$female$h2)) {
  knitr::kable(data$female$h2, digits = 4, caption = "Female h2 estimates (observed scale)")
}
```

## Exploratory Factor Analysis

### Male

```{r efa-male}
if (!is.null(data$male$efa_loadings)) {
  knitr::kable(data$male$efa_loadings, digits = 3, caption = "Male EFA loadings")
}
```

### Female

```{r efa-female}
if (!is.null(data$female$efa_loadings)) {
  knitr::kable(data$female$efa_loadings, digits = 3, caption = "Female EFA loadings")
}
```

## Confirmatory Factor Analysis

### Model Fit Comparison

```{r cfa-fit}
fit_all <- data.table()
for (sex in c("male", "female")) {
  if (!is.null(data[[sex]]$cfa_fit)) {
    tmp <- copy(data[[sex]]$cfa_fit)
    tmp[, sex := sex]
    fit_all <- rbind(fit_all, tmp, fill = TRUE)
  }
}
if (nrow(fit_all) > 0) {
  knitr::kable(fit_all, digits = 3, caption = "CFA model fit indices")
}
```

## Sex Comparison

```{r comparison}
if (!is.null(data$comparison) && nrow(data$comparison) > 0) {
  knitr::kable(data$comparison, digits = 3, caption = "Factor loading sex differences (unstandardized)")
} else {
  cat("No sex comparison data available.")
}
```

## Methods Note

- All h2 estimates are on the **observed scale** (no population prevalence applied).
- Effective sample size: Neff = 4 * n_cases * n_controls / (n_cases + n_controls).
- LDSC free intercept used to account for within-sex sample overlap.
- EFA and CFA applied to the same genetic covariance matrix (double-dipping caveat applies).
- Sex differences assessed via Wald Z-test on unstandardized loading differences from independent per-sex fits.
